### 2024-08-15 Paper/paper2.tex

    \begin{abstract}
		{\small
         Managers are often seen as drivers of firm decisions, strategies, and performance, and manager heterogeneity has been linked to employee wages, decisions of mergers and acquisitions, and financial performance. A significant goal in US health care is to improve quality while decreasing spending, and yet it is unclear whether hospital managers might contribute to this goal. In this paper, I estimate the effect of clinically trained executives on hospital quality. I leverage pay-for-performance initiatives that changed financial incentives on quality of care for hospitals in the US in 2012, and estimate the difference in readmission and mortality rates for nonprofit hospitals with and without clinically trained executives. I find that hospitals without clinically trained executives respond more to financial incentives than those without clinically trained executives. This difference is not driven by a change in patient complexity, but may be driven by preferences towards profit vs. patient welfare.  
		} 
	\end{abstract}

@@

    Executives are considered crucial drivers of a company's practices and performance, as they often set broad visions and implement strategies to achieve them. Executive heterogeneity has been linked to various decisions and outcomes of firms: characteristics such as gender, age, and background are correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, there is still much to learn about how firm behaviors and objectives can be shaped by differences in leadership.

    In this paper, I study how executive background affects firm behavior in the context of US hospitals. In the US, health care makes up almost 20\% of spending (\cite{CMS_National_Health_Expenditure}), and a major goal of policymakers is to improve the value of health care. Yet, little is known about how hospital managers affect health care value, and it is unclear whether existing findings on executive characteristics can be extrapolated to the health care context. This is because existing literature is generally limited to large, publicly traded firms, whereas a majority of US hospitals are nonprofit. Relatedly, health care executives have a particularly unique trade-off between profit and patient welfare, which varies from the types of firms studied in existing literature.

    A characteristic relevant to this context is whether or not hospital executives have clinical training. Physicians can bring a unique combination of clinical and administration expertise, and they have insights that can potentially benefit their employees and lead to better quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial, as physicians are not always trained in business management, which can lead to poor practices and lower quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how having clinically trained executives in hospital leadership affects quality of care. I scrape executive names and titles from publicly available Tax Form 990s, filled out by nonprofit firms in the US. Combining this with publicly available hospital characteristics, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is a very important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds on average (\cite{ASPE_2023}).

    The non-random selection of leaders makes it difficult to interpret any direct comparison of differing executive teams as causal. Therefore, I leverage pay-for-performance policies enacted in the US as an exogenous shock to hospital incentives, and compare the change in readmission and mortality rates of hospitals with and without clinically trained executives over the entire sample period. Under the assumption that the two types of teams would have had similar trends in readmission and mortality rates absent the incentive change, I find that hospitals without clinically trained trained executives respond more to financial incentives on quality than hospitals with clinical executives. That is, while both types of hospital leadership teams decrease readmission rates after the policy change, those without clinical executives decrease readmissions at a faster rate than those with clinical executives. Interestingly, this result is driven by hospitals with between one quarter and one half of their executive team having clinical training. Further, executives with clinical training in fields specific to the policy change respond more than other clinically trained executives.

    One potential reason for this differential response is that hospitals without clinically trained executives are more willing to participate in selective patient practices. To investigate this, I use patient case mix index as an outcome and estimate the difference in case mix index after the financial incentive change. I find no evidence that non-clinical executive teams differentially change patient complexity.

    Next, I investigate whether this differential response could be due to differences in objective functions. I consider hospital objectives as a function of profit and patient welfare, and allow weight placed on each to vary with executive team type. Hospitals that place more weight on profit will naturally respond more drastically to a policy that adds quality directly into the profit function. Thus, a potential reason for the difference in response is that non-clinical executive teams are more profit driven. I investigate this empirically by comparing both types of hospital leadership teams to for-profit hospitals, who have revealed that they are largely profit driven by their ownership type. This analysis reveals that non-clinical executive teams respond to the financial incentives similarly to for-profits, and hospitals with clinically trained executives act differently than for-profits.

    Finally, the question still remains whether clinically trained executives are a signal of underlying hospital preferences, or if they actively manage the hospital differently than their non-clinically trained counterparts. I analyze this by leveraging changes in hospital propensity to hire clinical executives to decompose the full effect of clinical leadership into a signaling effect (clinical executives signal underlying hospital preferences), or managing effect (clinical executives manage the hospital differently). First, I show that hospital executive changes are not correlated with the change to financial incentives. For the signaling effect, I compare hospitals who ever have a clinically trained executive to hospitals who never have a clinically trained executive. For the managing effect, I limit to hospitals who ever have a clinically trained executive and compare those who have a clinically trained executive when the policies were announced to those who have clinically trained executives at other times. I find that the difference in response is driven by a managing effect.

    Many researchers have documented a correlation between executive characteristics and firm performance, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Female board members are associated with better oversight and more women executives (\cite{matsa2011chipping}; \cite{adams2009women}), but are negatively correlated with firm value (\cite{ahern2012changing}). Having more female executives is correlated with female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions, while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}; \cite{adams2018director}; \cite{frydman2019rising}). Finally, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}).

@@

    I use several data sets to analyze the relationship between hospital executives with clinical training and response to financial incentives on quality. I construct a novel data set of hospital executives by scraping publicly available tax forms, and merge this data to publicly available data sets containing hospital characteristics: the American Hospital Association Survey (AHA), the Centers for Medicare and Medicaid Services (CMS) Hospital Compare data and Case Mix Index files, and Healthcare Cost Report Information System (HCRIS) data. Ultimately, I employ this hospital-level data spanning from 2010-2014, which contains executive team characteristics, mortality and readmission rates as measures of quality, and other characteristics. A detailed description of the construction of this data can be found in Appendix \ref{appendixdata}.

@@

    For each hospital in the tax form data, I seek matches in the AHA data based on hospital name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optimal character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. While I extract text spanning 2009-2015, I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 852 nonprofit hospitals that are matched to AHA data, and with complete text information in each year.

    Finally, I seek to (1) verify clinical training of those who are identified as doctors, and (2) obtain additional information about clinically trained executives. Thus, I match each name with the National Plan and Provider Enumeration System (NPPES) database of all registered physicians. I present a table of means for various characteristics of doctor executives in the data in Table \ref{doc_sumstats}. In total, there are 813 clinically trained executives. On average, they are 52 years old, and 11\% of clinically trained executives are female. While it is more likely they are a Chief Medical Officer than a Chief Executive Officer, still less than a third are CMOs. The most common specialty is internal medicine.

@@

    For the purposes of this paper, I focus on executive teams as a whole instead of specific positions. The main reason for this is because the organization of teams varies so drastically across hospitals. It is not uncommon for two hospitals to have the same position but the tasks performed are drastically different, or to have two distinct positions that essentially perform the same tasks. I create various hospital-level characterizations based on their executive team, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer.

@@

    The purpose of this paper is to investigate hospital behavior in the midst of changes to financial incentives that change how hospitals receive payment for quality of care. Two programs were passed as part of the Affordable Care Act that focused on pay-for-performance incentives for hospitals: the Hospital Readmissions Reduction Program (HRRP), and the Hospital Value Based Purchasing Program (HVBP). HRRP focused on penalizing hospitals with poor quality, and HVBP focused on rewarding hospitals with high quality or show improvement in quality.

@@

    Next, I plot weighted average readmission, mortality rates, and patient case mix index across time for different hospital types in Figure \ref{fig:outcomes_graph}. The effective start year of pay-for-performance incentives is shown as a dotted line in 2012. Average readmission rates trend similarly prior to 2012, and all hospital types exhibit a decrease in raw readmission rates after 2012. However, there is a faster rate of decrease for hospitals who have a clinically trained executive at some point, or who never have a clinically trained executive. When considering mortality, there is not a drastic change in 2012 for any hospital type, but hospitals who always have a clinical executive continue on an increasing trend in mortality where the other hospitals types remain more flat. Finally, hospitals that never have clinical executives consistently have less complex patients, but there is no notable change in raw case mix index over time.

@@

    \section{Effect of Executive Clinical Training on Response to Quality Incentives}\label{sec:clinical}

    First, I test empirically whether having clinical training matters in how hospitals change readmission and mortality rates in response to the pay-for-performance policies effective in 2012. For a clean comparison, I limit the sample to hospitals who always have clinical executives or never have clinical executives. That is, these hospitals do not change their propensity to hire a clinical executive during the sample period. \footnote{An important assumption for this analysis is that this sample selection is not correlated with the program enactment. I investigate this assumption in Section \ref{sec:changes}, and find no evidence that hospitals are more or less likely to change their propensity to hire a clinical executive because of the pay-for-performance programs. }

    I estimate a two-way-fixed-effects specification, where the key independent variable is the interaction between having clinically trained executives and post 2012, when the programs began. Importantly, I am not estimating the effect of being penalized. All hospitals had incentive to improve quality in response to the policy change, as penalties and payments are determined by comparing quality among hospitals. Thus, I leverage the timing of the policy change, rather than the policies themselves. However, because of potential differences in quality trends between clinical and non-clinical hospital executive teams even before the policies, I employ a synthetic difference-in-differences estimation strategy (\cite{arkhangelsky2021synthetic}). That is, hospitals in the control group with similar trends in readmission, mortality, and baseline case mix before the programs are more heavily weighted, as well as time periods that balance pre- and post-program outcomes for hospitals without clinical executives. Below is the unweighted specification for clarity, where the identification assumptions are conceptually the same for synthetic difference-in-differences:

    \begin{equation}
    \label{eq:clinical}
    y_{ht} = \beta \text{ (clinical training x post 2012)}_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht},
    \end{equation}

    \noindent where $y_{ht}$ is one of the outcome variables capturing readmission or mortality rates discussed in Section \ref{sec:hrrp}, and $\gamma_h$ and $\delta_t$ are hospital and time fixed effects, respectively.

    In the unweighted two-way-fixed-effects specification, we have reason to believe that the parallel trends assumption may not hold due to observed differences in characteristics other than the leadership team, as discussed in Section \ref{sec:data}. Once incorporating the weights used in the synthetic differences in differences method, we ensure parallel pre-trends in the outcome variables. Thus, I assume that under the weighted clinical and non-clinical hospital composition, trends would have continued to be parallel absent the incentive change in 2012. This includes an implicit assumption that no other event occurred in 2012 correlated with both the outcomes and composition of leadership team. Finally, I assume that there is no anticipation of the policy change before 2012. The rules of the policies were announced in October of 2011, so 2012 was the first full year that hospitals could meaningfully respond to the policies.

    The estimated difference in readmission and mortality rates between hospitals with and without clinical executives are shown in Figure \ref{fig:clinicalsynthdid}. Estimate $\beta$ is represented in the difference in slope after 2012 when pay-for-performance incentives began. Panel \ref{fig:read_synth_clinical} shows the estimated difference in readmission rates, and reveals that non-clinical teams decrease readmissions by .3 (standard error .13) more than clinical teams. This is a relatively small magnitude given an average readmission rate of 21, but reveals a meaningful difference in behavior given that both hospital types significantly decrease readmissions. In panel \ref{fig:mort_synth_clinical}, I show the estimated difference in mortality. While the magnitude of the difference is similar to that for readmission rates, mortality is a noisy measure, and I cannot rule out null effects for mortality. In Appendix \ref{app:condition}, I show that these results are primarily driven by pneumonia patients.

@@

    Additionally, I estimate the intensive margin of clinical leadership: whether the fraction of clinical executives on the team affects quality. I categorize hospitals with clinically trained executives into two bins: above the median and below median fraction of clinically trained executives. Let the median fraction be $\eta$, which in this case is .2. Then, I estimate the synthetic difference-in-differences version of the following specifications:

@@

    \noindent where $\rho$ is the fraction of clinical executives on hospital $h$'s executive team, and $\gamma$ and $\delta$ are fixed effects as defined in Equation \ref{eq:clinical}.

    I present the results of this estimation with readmission rates as the outcome in Figure \ref{fig:cont_clinicalsynthdid}. In panel \ref{fig:belowmed_read_synth_clinical}, the estimation compares executive teams with no clinical training to those with up to 25\% of the team having clinical training. The average treatment effect is .04 and not statistically significant. Thus, there is no significant difference in how these types of teams respond to the incentive change. In panel \ref{fig:abovemed_read_synth_clinical}, I compare hospitals with no clinically trained executives over the course of the sample to hospitals with over a quarter of their team having clinical training. I find that hospitals without clinically trained executives decrease readmissions by .64 (standard error .33) more than hospitals with clinically trained executives. Thus, the readmission rate response is fully driven by having a higher proportion of clinical executives relative to no clinical executives.

@@

    Any potential difference in mortality rate response also seems to be driven by executive teams with over a quarter of the executives having clinical training, as shown in Figure \ref{fig:cont_clinicalsynthdid_mort}. However, the standard errors of this estimation are high, potentially due to the noisiness in the mortality measure. The magnitude of the estimate in panel $\ref{fig:belowmed_mort_synth_clinical}$ is .07, so I can rule out differential responses by teams with less than a quarter of clinical executives. Taking the readmission and mortality rate results together, the effect of having clinical leadership on quality after an incentive change depends on having a sufficient proportion of the team with clinical training.

@@

    \section{Selective Patient Practices}

    \begin{figure}[ht!]
        \centering
        \caption{Effect of Clinical Executive on Cases Mix Index}
        \includegraphics[width=0.5\textwidth]{Objects/cmi_md_nomd_synth_graph.pdf}
        \label{fig:main_cmi_clinical}
    \end{figure}

    \section{Internal Medicine vs. Other Specialties}

    While most doctors still take a hippocratic oath to provide care with integrity and compassion, doctors with training in certain specialties may be more equipped to improve care for patients with the particular conditions relevant in this context. Specifically, a pulmonologist typically treats pneumonia, and a cardiologist typically treats heart failure and heart attack patients, which are the three conditions considered in the HRRP penalties. Both of these specialties are under the umbrella of internal medicine, which is the most granular specialty type in my data.

@@

    Another hypothesis that could explain the differential response due to clinical leadership is how profit-driven a clinician is compared to other training backgrounds. To demonstrate this, consider hospital behavior under the enactment of pay-for-performance policies, where hospitals face a trade-off between profit and societal benefit. Pay-for-performance incentives essentially incorporate quality directly into a hospital's profit function (\cite{dranove2011health}), whereas a fee-for-service system only incentivizes quantity. A hospital that places all value on profit will only care about quality when it enters into the profit function, and therefore will respond more drastically to pay-for-performance than a hospital who already cared about quality through societal benefit before the incentives.\footnote{I present the theoretical model and mathematical results in Appendix \ref{sec:model}.} Therefore, observing hospital responses to pay-for-performance incentives reveals information about their underlying preferences.

    I merge all general for-profit hospitals in the AHA data to the sample of nonprofit hospitals. Then, I compare readmission and mortality rates of always clinical executives and never clinical executives to for-profit hospitals after the incentive change. Ideally, I would also have leadership heterogeneity information for for-profit hospitals, but this data does not exist publicly. However, on average, for-profit hospitals have revealed through their ownership status that they more heavily weigh profit. Thus, comparing different executive team behavior to for-profit behavior helps reveal how profit driven certain leadership teams are. Thus, I estimate

@@

    Next, I assess whether this assumption is reasonable by analyzing how executive team changes are occurring over time for certain groups of hospitals. I estimate:

@@

     \subsection{Estimation and Findings}

>>>>

    \begin{abstract}
		{\small
         Managers are often seen as drivers of firm decisions, strategies, and performance. Further, manager characteristics have been linked to employee wages, decisions of mergers and acquisitions, and financial performance. Health care policies in the US often focus on improving quality of care without increasing costs, and yet it is unclear whether hospital managers might contribute to this goal. In this paper, I estimate the differential response to such quality incentives by different types of nonprofit hospital executives: those with and without a clinical training background. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. I provide evidence that this may be due to differing preferences for profit vs. patient welfare. Thus, incentivizing clinically trained executives in hospitals could be an effective method to improve value in the US health care system. 
		} 
	\end{abstract}

@@

    In corporate environments, executives are considered crucial drivers of a company's practices and performance, as they are heavily involved in the culture, visions, and strategies of the firm. Executive heterogeneity has been linked to various actions and outcomes of firms; characteristics such as gender, age, and background are correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, there are still settings and relevant characteristics for which the relationship between management and firm behavior is not well understood.

    In this paper, I study whether type of executive training affects firm behavior in the context of US hospitals. In the US, health care makes up almost 20\% of spending (\cite{CMS_National_Health_Expenditure}), and a major goal of policymakers is to improve the quality of health care without increasing costs. Yet, little is known about how hospital managers affect health care value, and it is unclear whether existing findings on executive characteristics can be extrapolated to the health care, and particularly hospital, context. Existing literature is generally limited to large, publicly traded firms, whereas a majority of US hospitals are nonprofit with presumably different objectives. In particular, health care executives have a particularly unique trade-off between profit and patient welfare, which varies from the types of firms studied in existing literature.

    A characteristic relevant to this context is whether or not hospital executives have clinical training. Physicians can bring a unique combination of clinical and administration expertise, and they have insights that can potentially benefit their employees and lead to better quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial, as physicians are not always trained in business management, which can lead to poor practices and lower quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how having clinically trained executives in hospital leadership affects hospital outcomes. I scrape executive names and titles of nonprofit hospitals in the US from publicly available Tax Form 990s. Combining this with other public hospital data, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is an important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds on average (\cite{ASPE_2023}).

    The non-random selection of leaders makes it difficult to interpret any direct comparison of differing executive teams as causal. Therefore, I leverage pay-for-performance policies enacted in the US as an exogenous shock to hospital incentives on quality, and compare the change in readmission and mortality rates of hospitals with and without clinically trained executives. Under the assumption that the two types of teams would have had similar trends in readmission and mortality rates absent the incentive change, I find that hospitals without clinically trained trained executives respond more to financial incentives on quality than hospitals with clinical executives. That is, while both types of hospital leadership teams decrease readmission rates after the policy change, those without clinical executives decrease readmissions at a faster rate than those with clinical executives. I find no evidence of selective patient practices driving this result.

    I investigate whether the result might be explained by differing preferences on the trade-off between profit and patient welfare, where I model these hospital objectives and allow weight placed on each to vary with executive team type. Hospitals that place more weight on profit will naturally respond more drastically to a policy that adds quality directly into the profit function. Thus, a potential reason for the difference in response is that non-clinical executive teams are more profit driven. I investigate this empirically by comparing the response of hospitals with and without clinically trained executives to the response of for-profit hospitals, who have revealed that they are more profit driven by their ownership type. I find that nonprofit hospitals without clinically trained executives respond to the financial incentives on quality similarly to for-profits, while nonprofits without clinically trained executives respond differently than for-profits. For-profits and non-clinical executive teams change quality more in response to the incentive change than clinically trained executive teams.

    Finally, the question remains whether clinically trained executives are a signal of unchanging underlying hospital preferences, or if they actively manage the hospital differently, effectively changing hospital objectives. I analyze this by leveraging changes in hospital propensity to hire clinical executives to decompose the full effect of clinical leadership into a signaling effect (clinical executives signal underlying hospital preferences), or managing effect (clinical executives manage the hospital differently). First, I show that hospital executive changes are not correlated with the change to financial incentives. For the signaling effect, I compare hospitals who ever have a clinically trained executive to hospitals who never have a clinically trained executive. For the managing effect, I limit to hospitals who ever have a clinically trained executive and compare those who have a clinically trained executive when the policies were announced to those who have clinically trained executives at other times. I find that the difference in response is driven by clinically trained executives managing the hospital differently.

    Many researchers have documented a correlation between executive characteristics and firm actions, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Having more female executives is correlated with female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions, while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}; \cite{adams2018director}; \cite{frydman2019rising}); older CEOs are associated with measures of corporate culture (\cite{graham2022corporate}). Finally, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}).

@@

    I use several data sets to analyze the relationship between executives with clinical training and hospital response to financial incentives on quality. I construct novel information on individual hospital executives by scraping publicly available tax forms, and merge this data to publicly available data sets containing other hospital characteristics: the American Hospital Association Survey (AHA), the Centers for Medicare and Medicaid Services (CMS) Hospital Compare data and Case Mix Index files, and Healthcare Cost Report Information System (HCRIS) data. Ultimately, I employ this hospital-level data spanning from 2010-2014, which contains executive team characteristics, mortality and readmission rates as measures of quality, and other characteristics. In this section, I describe the institutional setting and data collection process. A more detailed description of the construction of this data can be found in Appendix \ref{appendixdata}.

@@

    For each hospital in the tax form data, I merge to hospital matches in the AHA data based on name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optimal character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. I extract text spanning 2009-2015, but I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 852 nonprofit hospitals that are matched to AHA data, and with complete text information in each year.

    Finally, I verify clinical training of those who are identified as doctors, and obtain additional information about clinically trained executives by matching names to the National Plan and Provider Enumeration System (NPPES) database of all registered physicians. I present a table of means for various characteristics of doctor executives in the data in Table \ref{doc_sumstats}. In total, there are 813 clinically trained executives. On average, they are 52 years old, and 11\% of clinically trained executives are female. While it is more likely they are a Chief Medical Officer than a Chief Executive Officer, still less than a third are CMOs. The most common specialty is internal medicine.

@@

    I focus on executive teams as a whole instead of specific positions because the organization of teams varies so drastically across hospitals. It is not uncommon for two hospitals to have the same position but the tasks performed are drastically different, or to have two distinct positions that essentially perform the same tasks. I create various hospital-level characterizations based on their executive team, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer.

@@

    The purpose of this paper is to investigate hospital behavior in the midst of financial incentives that change how hospitals receive payment for quality of care. Two programs were passed as part of the Affordable Care Act that focused on pay-for-performance incentives for hospitals: the Hospital Readmissions Reduction Program (HRRP), and the Hospital Value Based Purchasing Program (HVBP). HRRP focused on penalizing hospitals with poor quality, and HVBP focused on rewarding hospitals with high quality or show improvement in quality.

@@

    Next, I plot weighted average readmission rates, mortality rates, and patient case mix index across time for different hospital types in Figure \ref{fig:outcomes_graph}. The effective start year of pay-for-performance incentives is shown as a dotted line in 2012. Average readmission rates trend similarly prior to 2012, and all hospital types exhibit a decrease in raw readmission rates after 2012. However, there is a faster rate of decrease for hospitals who have a clinically trained executive at some point, or who never have a clinically trained executive. When considering mortality, there is not a drastic change in 2012 for any hospital type, but hospitals who always have a clinical executive continue on an increasing trend in mortality where the other hospitals types remain more flat. Finally, hospitals that never have clinical executives consistently have less complex patients, but there is no observable change in raw case mix index over time.

@@

    \section{Effect of Clinically Trained Executives on Response to Financial Incentives on Quality}\label{sec:clinical}

    I test empirically whether having clinically trained executives matters in how hospitals respond to the pay-for-performance policies effective in 2012. The quality metrics targeted by the programs are readmission and mortality rates, so I use these as outcomes. I limit the sample to hospitals who always have clinical executives or never have clinical executives. That is, the hospitals in this analysis do not change their propensity to hire a clinical executive during the sample period.\footnote{An important assumption for this analysis is that this sample selection is not correlated with the program enactment. I investigate this assumption in Section \ref{sec:changes}, and find no evidence that hospitals are more or less likely to change their propensity to hire a clinical executive because of the pay-for-performance programs. }

    I ultimately employ a synthetic difference-in-differences estimation strategy that mirrors two-way-fixed effects, where the key independent variable is the interaction between having clinically trained executives and being in years after pay-for-performance took effect. All hospitals had incentive to improve quality in response to the policy change, as penalties and payments are determined by comparing quality among hospitals. Thus, I leverage the timing of the policy change, rather than the policies themselves. One may be concerned about potential differences in quality trends between clinical and non-clinical executive team hospitals leading to a violation in the parallel trends assumption. This is why I implement a synthetic difference-in-differences estimation strategy (\cite{arkhangelsky2021synthetic}). That is, non-clinical team hospitals with trends similar to clinical team hospitals in readmission, mortality, and baseline case mix before the programs are more heavily weighted, as well as time periods that balance pre- and post-program outcomes for non-clinical teams. I present the unweighted two-way-fixed-effects specification equation for clarity, and the identification assumptions are conceptually similar for synthetic difference-in-differences.

    \begin{equation}
    \label{eq:clinical}
    y_{ht} = \beta \text{ (clinical training x post 2012)}_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation}

    \noindent The variable $y_{ht}$ is one of the outcome variables capturing readmission or mortality rates discussed in Section \ref{sec:hrrp}, and $\gamma_h$ and $\delta_t$ are hospital and time fixed effects, respectively.

    Once incorporating the weights used in the synthetic differences-in-differences method, we ensure parallel pre-trends in the outcome variables. Thus, I assume that under the weighted clinical and non-clinical hospital composition, trends would have continued to be parallel absent the incentive change in 2012. This includes an implicit assumption that no other event occurred in 2012 correlated with both the outcomes and composition of leadership team. Finally, I assume that there is no anticipation of the policy change before 2012. The rules of the policies were announced in October of 2011, so 2012 was the first full year that hospitals could meaningfully respond to the policies.

    The estimated difference in readmission and mortality rates between hospitals with and without clinical executives are shown in Figure \ref{fig:clinicalsynthdid}. Estimate $\beta$ is represented in the difference in slope after 2012 when pay-for-performance incentives began. Panel \ref{fig:read_synth_clinical} shows the estimated difference in readmission rates, and reveals that non-clinical teams decrease readmissions by .3 (standard error .13) more than clinical teams. This is a relatively small magnitude given an average readmission rate of 21, but reveals a meaningful difference in behavior given that both hospital types significantly decrease readmissions. In Appendix \ref{app:condition}, I show that these results are primarily driven by pneumonia patients.

    In panel \ref{fig:mort_synth_clinical}, I show the estimated difference in mortality. While the magnitude of the difference is similar to that for readmission rates (relative to the mean), mortality is a noisy measure, and I cannot rule out null effects. I investigate whether the estimated difference in readmission rate response could be due to differential participation in selective patient practices in Appendix \ref{app:casemix}, and find that this is not the case.

@@

    Additionally, I estimate the intensive margin of clinical leadership: whether the fraction of clinically trained executives matters for response to financial incentive changes. I categorize hospitals with clinically trained executives into two bins: above the median and below median fraction of clinically trained executives. Again, I present the unweighted version of the synthetic difference-in-differences that I ultimately employ:

@@

    \noindent where $\rho$ is the fraction of clinical executives on hospital $h$'s executive team, $\eta$ is the median fraction of clinically trained executives, which is .2 in this sample, and $\gamma$ and $\delta$ are hospital and time fixed effects.

    I present the results of this estimation with readmission rates as the outcome in Figure \ref{fig:cont_clinicalsynthdid}. In panel \ref{fig:belowmed_read_synth_clinical}, I compare executive teams with no clinical training to those with a fraction of clinically trained executives up to the median value, 20\%. With an average treatment effect of .21 (standard error .31), hospitals having this range of clinically trained executives are not driving the main finding. However, in panel \ref{fig:abovemed_read_synth_clinical}, I compare hospitals with no clinically trained executives to hospitals with a fraction of clinically trained executives over the median of 20\%. I find that hospitals without clinically trained executives decrease readmissions by .6 (standard error .14) more than hospitals with clinically trained executives. Thus, the readmission rate response is driven by having a higher proportion of clinical executives relative to no clinical executives.

@@

    I present the estimation results for mortality as an outcome in Figure \ref{fig:cont_clinicalsynthdid_mort}. Unsurprisingly given the noisiness in the main findings for mortality, it is unclear whether there is any intensive margin effect for mortality. Both average treatment effect magnitudes are lower than that for the main result with high standard errors. Thus, the intensive margin does not seem to matter as much as it does for decreasing readmission rates.

@@

@@

    One hypothesis that could explain the differential response due to clinical leadership is how profit-driven a clinician is compared to other training backgrounds. To demonstrate this, consider hospital behavior under the enactment of pay-for-performance policies, where hospitals face a trade-off between profit and societal/patient benefit. Pay-for-performance incentives essentially incorporate quality directly into a hospital's profit function (\cite{dranove2011health}), whereas a fee-for-service system only incentivizes quantity. A hospital that places all value on profit will only care about quality when it enters into the profit function, and therefore will respond more drastically to pay-for-performance than a hospital who already cared about quality through societal benefit before the incentives.

    $$\max_{\theta}\hspace{2mm}\alpha\pi(\theta) + (1-\alpha) u(\theta),$$

    \noindent where $\pi(\theta)$ is a profit function depending on $\theta$, and the relationship between profit and $\theta$ depends on pay-for-performance incentives. Additionally, $u(.)$ is extra utility gained from quality, which is increasing and concave in $\theta$, and $\alpha\in[0,1]$ captures the weight placed on profit vs. societal benefit. There is an implicit stay-open condition throughout the model. Taking the first order condition yields

    $$(1-\alpha)u'(\theta) = -\alpha \pi'(\theta),$$

    $$\frac{du'(\theta)}{d\alpha} = \frac{-1}{(1-\alpha)^2}\pi'(\theta)$$

    First, assume the hospital is in a strictly fee-for-service setting, where quality benefits patients, but does not directly provide profit to the hospital through reimbursements. Then, assuming there is a positive cost to increasing quality, $\pi(\theta)$ is decreasing in $\theta$. Then $\frac{du'(\theta)}{d\alpha} > 0$, revealing that in a fee-for-service setting, the more weight placed on profit, the lower quality of care chosen by the hospital.

    Second, assume the hospital is in a strictly pay-for-performance setting, where quality benefits patients and directly increases profit. Then, assuming the marginal financial benefit of increasing quality is greater than the marginal financial cost, $\pi(\theta)$ is increasing in $\theta$. Then $\frac{du'(\theta)}{d\alpha} < 0$, revealing that in a pay-for-performance setting, the more weight placed on profit, the higher quality of care chosen by the hospital. I combine the results found in each scenario into one response function that depends on $\alpha$, where $\theta_{p4p}$ is quality under pay-for-performance incentives and $\theta_{ffs}$ is quality under fee-for-service only. That is,

    \begin{align*}
        \frac{d\Delta\theta}{d\alpha}&=\frac{d(\theta_{p4p}-\theta_{ffs})}{d\alpha}\\
        &=\frac{d\theta_{p4p}}{d\alpha}-\frac{d\theta_{ffs}}{d\alpha}\\
        &> 0.
    \end{align*}

    Hence, this model predicts that change in quality depends on how much weight the hospital places on profit vs. societal benefit. Particularly, hospitals with more weight on profit respond more than hospitals with more weight placed on societal benefit. In reality, $\alpha$ is unknown, except in the case of for-profits, who have revealed that they have a high $\alpha$ by their ownership status. Clinically trained executives could also affect this $\alpha$, as doctors commonly take a hippocratic oath to put patients first.

    \subsection{Estimation}

    To investigate this hypothesis empirically, I merge all general for-profit hospitals in the AHA data to the sample of nonprofit hospitals. Then, I compare readmission and mortality rates of always clinical executives and never clinical executives to for-profit hospitals after the incentive change. Ideally, I would also have leadership heterogeneity information for for-profit hospitals, but this data does not exist publicly. However, on average, for-profit hospitals have revealed through their ownership status that they more heavily weigh profit. Thus, comparing different executive team behavior to for-profit behavior helps reveal how profit driven certain leadership teams are. Thus, I estimate the weighted synthetic difference-in-differences version of

@@

    Next, I assess whether this assumption is reasonable by analyzing how executive team changes are occurring over time for certain groups of hospitals. I estimate the following two-way-fixed-effects specification:

@@

     \subsection{Estimation and Findings}\label{sec:decomp}
